Climate-driven phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-07-27 agujournaltemplate.tex
Results checked and Open Research Statement added

% Topic Overview
Phytoplankton are microscopic algae that form the foundation of the ocean food web. Because they respond quickly to environmental changes, studying them can show how marine ecosystems are affected by climate.
% Paper Overview
This study looked at how phytoplankton communities in the Cariaco Basin, off the coast of Venezuela, changed from 1995 to 2017. We compared microscopy cell counts and monthly water samples with weather data and two major climate indicators: the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both linked to local wind and currents.
% Findings Summary
The results showed a major shift in the phytoplankton community in 2004, linked to a decrease in upwelling, with conditions partially returning to the previous state by 2014. Interestingly, species diversity declined between 1998 and 2009, although upwelling had not yet changed. Further analysis showed that the AMO index, along with sea surface temperature, MEI v.2 and nitrate levels in the water column, were the strongest predictors of changes in the phytoplankton community.
% Key Takeaways
These findings contribute to a broader understanding of how climate variability shapes marine phytoplankton communities and underscore the importance of sustained long-term observations in detecting and interpreting ecological responses to climate change.

@@

\subsection{In-situ biological and environmental data}
    The CARIACO Ocean Time Series Program collected biological and environmental data, which we used here to examine the connections between environmental and ecological change in the region. Samples were collected on monthly research cruises at a station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.67}W) \cite{muller-karger_scientific_2019}.
    The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, along with other deployments and samples not utilized in this work. Details about the sampling procedures, analytical methods, data quality control, and inter-calibration procedures have already been described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.

@@

    All in situ data, including the Niskin bottle, CTD, and phytoplankton taxonomy, were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047}. 
    From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps (no more than three months) in coverage. In total, we used a complete set of 254 monthly data points from 1995 to 2017, for which all variables were available.

\subsection{Meteorological and climate data}
    %from Taylor et al. 2012
    In addition to oceanographic environmental data, we also examine meteorological and climatic data that are known to influence ocean conditions and marine ecosystems. Zonal wind (\textit{u}, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters \cite{rueda-roa_southern_2013}. We extracted the variable u10, the \qty{10}{\meter} eastward component of wind speed from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels}. For simplicity, we refer to this variable as "wind speed" throughout the article. The data for precipitation and evaporation are also extracted from the ERA5 dataset for the aforementioned grid cell. Evaporation provides an estimate of the accumulated amount of water that has evaporated from Earth's surface, while precipitation is an estimate of total precipitation, comprising all water that falls on Earth's surface as rain or snow.

    In addition to the wind data, we examined two climatic indices, each of which provides information on the basin-scale climatic conditions and, consequently, the broad-scale forcing of marine ecosystems. The monthly Atlantic Multidecadal Oscillation (AMO) index data (10-year low-pass), which represent area-averaged, detrended low-pass filtered North Atlantic SST anomalies (\ang{0}N-\ang{60}N), were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo}.

    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure, SLP; sea surface temperature, SST; zonal and meridional components of surface wind; and outgoing long-wave radiation, OLR) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W), were obtained from \url{https://psl.noaa.gov/enso/mei/}. In our analysis, bi-monthly values were aligned with the tail month (December-January is treated as the value for January).

    \subsubsection{Data processing}
    % Niskin, CTD and Phytoplankton data processing
    Data were processed and analyzed using the R programming language (version 4.3.3) \cite{r_core_team_r_2024} with data pipelines created using tidyverse packages \cite{wickham_welcome_2019}. All the codes we produced for data processing and data analysis are publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.
    For all discrete depth measurements, the data were interpolated using the "oce" package \cite{kelley_oce_2023} with the adaptive "unesco" algorithm defined by the U.S. National Oceanographic Data Center \cite{johnson2006world}. The interpolated values were depth integrated to 100 meters and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 \meter for further analysis. Additionally, sea surface temperature (SST) was calculated from CTD temperature measurements averaged over the top \qty{10}{\meter}. For phytoplankton microscopy data, individually identified taxonomic units (species or genus) were interpolated and integrated across depth, and finally grouped by genus via the sum of interpolated counts.

\subsection{Data analysis}    
    In the analysis, we set out to find if there is a structure in the distribution and temporal evolution of diversity metrics and the community composition and to identify potential bottom-up drivers of this change. 
    To provide an overview of the available data and highlight interannual trends and changes in the time series data, we calculated annual anomalies for relevant variables as normalized z scores ($z = \frac{(x – \mu)}{\sigma}$, where \mu is the mean across all available data points). Z scores from 1995, 2016, and 2017 were not shown because insufficient coverage (i.e. more than six missing monthly samples) in these years prevented us from producing reliable yearly estimates.

@@

    \subsubsection{Assessing Environmental Drivers of Phytoplankton Dynamics}
    To quantify the effects of bottom-up drivers in the Cariaco Basin on shifts in the phytoplankton community, we employed a Gradient Forest analysis. Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}. Gradient Forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor. The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community turnover. Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}. We used monthly genus level abundances integrated over the top \qty{100}{\meter} as indicator variables and limited the data to only those genera that were counted more than 10 times over the entire time series. This produced a data set of 193 monthly samples and 89 observed genera abundances, where all predictor variables were also available. The genus abundances were transformed by the natural logarithm plus one prior to model runs. The drivers/predictor variables considered in the analysis were the AMO and MEI v.2 indices, wind speed, evaporation, precipitation, \qty{21}{\celsius} isotherm, SST, salinity, and nutrient concentrations ($\mathrm{[NO_3]}$, $\mathrm{[PO_4]}$ and $\mathrm{[SiO_4]}$). All predictor variables are in some way related to the local climate and upwelling system and could be potential bottom-up drivers of community change, some more directly (e.g. the isotherm and nutrient concentrations) and some more indirectly, such as the climate indices and meteorological data. 
    The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology accounts for by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}. Gradient forest runs were performed with 1500 bootstrapped tress and a correlation threshold of 0.5.
    We show a model run without any time lags in the supplementary material (Figure \ref{fig:sup:GFoutput_nolags}). To consider the temporal lag in effect, particularly of climate indices, we tested for an improved predictive power of time lags of up to 6 months for all climate variables, as well as 3-month lags for in situ measurements in exploratory runs. We chose time lags that scored highest in the goodness-of-fit statistics for the final model run (Table \ref{tab:sup:GFlagtests} in Appendix). All the code and data used in this analysis and to create the plots are publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.

\caption{Data and corresponding units used as predictor variables for the gradient forest analysis.}
\centering
\resizebox{\textwidth}{!}{\begin{tabular}[t]{rrl}
\toprule
Variable Name & Units  & Data source \\
\midrule
MEI v.2 & unitless & Climate index provided by NOAA \\
AMO & unitless & Climate index provided by NSF NCAR \\
Wind speed & \qty{}{\meter/\second} & ERA5 model reanalysis, negative 10 m u-component of wind speed \\
Precipitation & \qty{}{\meter} & ERA5 model reanalysis, total precipitation\\
Evaporation & \qty{}{\meter} & ERA5 model reanalysis, evaporation \\
SST & \qty{}{\celsius} & CTD, Average temperature measured down to \qty{10}{\meter} depth\\
\qty{21}{\celsius} isotherm & \qty{}{\meter} & CTD, depth at which temperature reaches 21°C\\
Salinity & PSU & Niskin, interpolated down to \qty{100}{\meter} depth \\
$\mathrm{NO}_3$  & µM & Niskin, interpolated down to \qty{100}{\meter} depth \\
$\mathrm{PO}_4$  & µM & Niskin, interpolated down to \qty{100}{\meter} depth \\
$\mathrm{SiO}_4$  & µM & Niskin, interpolated down to \qty{100}{\meter} depth \\
\end{tabular}}
\label{tab:DataSources}
\end{table}

@@

\subsection{Interannual changes in the environment and ecosystem}
% Figure 2
Coordinated, interannual changes occur in normalized yearly means of all environmental variables considered, the logarithmic transformation of chlorophyll \textit{a}, and phytoplankton cell counts per functional group (Figure \ref{fig:zscore}). The patterns emerging from wind speed, evaporation, precipitation, SST, and \qty{21}{\celsius} isotherm depth clearly show the years of positive anomalies of wind (blue) and negative anomalies for SST (red) and upwelling (red, i.e., shallower depth of the isotherm). These patterns are reflected in the mode of the AMO, where a positive anomaly correlates with weak upwelling regimes (see also Figure \ref{fig:sup:correlation} in Appendix). The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series but show differing interannual oscillations. Annual mean anomalies of nitrate and phosphate appear to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some variability in mean phosphate concentration. Silicate does not follow a clear pattern but shows an elevated mean concentration between 2008 and 2013, which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion due to reduced abundances of diatoms.  <I think the most obvious pattern in Fig. 2 has not yet been discussed: Chl and phyto are high in 1996-2004, then low in 2004-2013, followed by a switch in 2014-2015. Diversity is high when chl/phyto are low. I think you'll want to describe these patterns as they are interesting.>

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/Figure2_ZScores_v2.pdf}
\caption{Time series showing annual z-scores of selected environmental, climatological, and ecological variables. Red indicates negative annual anomalies and blue indicates positive annual anomalies. Deeper shades indicate larger anomalies. The years 1995 (n=2) and 2016 (n=5), 2017 (n=1) were removed due to insufficient monthly data to generate robust annual anomalies. Additionally we marked years with two or three and four or more missing monthly values. Diversity indices were calculated based on microscopy counts at the genus level.}
\label{fig:zscore}
\end{figure}

% Figure 3
Figure \ref{fig:divts} shows the depth distribution and dynamics of chlorophyll \textit{a} and community diversity indices, also in relation to rainy and upwelling seasons. The bulk of the chlorophyll \textit{a} ($\sim$ 80\,\%) was measured in the upper 50 meters of the water column, with approximately 15\,\% measured between 50 and 75 meters. Chlorophyll \textit{a} exhibited marked seasonality, with the highest biomass in the upper water column during the early upwelling months (December–February, Figure \ref{fig:divts}$b$). The yearly mean chlorophyll \textit{a} biomass showed the highest variability within the upper 25 meters. In some years, particularly in 2000, 2005, 2007, 2008, and 2013, the concentration in this layer was lower than in the 25–50 meter depth range, suggesting a deep chlorophyll maximum in those years (Figure \ref{fig:divts}$a$). In general, yearly mean anomalies within the upper 75 meters appear to be high during the period from 1996 to 2004 and only reduced in 2012. The year 2013 was an outlier, particularly for the upper 25 meters, as the missing monthly sampling from the winter months (January and February) skewed the value downwards. Nevertheless, the monthly measurements for the remainder of the year 2013 show a markedly lower cholorophyll (65\,\% of the mean over the entire time series for these months), which recovers in 2014. Chlorophyll \textit{a} biomass below 50 meters depth showed a relatively stable distribution throughout the year and across the time series. In fall, during the rainy season, mean chlorophyll \textit{a} biomass in the upper 25 meters is markedly reduced, while the 50–75 meter depth interval showed a slight increase, indicating a shift in the chlorophyll \textit{a} maximum to lower depths under weak upwelling conditions.

The calculated genus diversity from the microscopy cell counts generally follows the dynamics of chlorophyll \textit{a} biomass, with maximum values observed during the upwelling seasons in winter. Genus richness is consistently the highest diversity index in the top \qty{25}{meters} depth interval, and shows also the highest variability over the time series. Across the time series, we observe maximum richness at the start of the time series (1996-1997), with a subsequent reduction in diversity across all depths, with a recovery after 2011. The  reduction in diversity following the start of the time series does not follow the pattern shown by chlorophyll \textit{a}, which remains relatively stable across the top \qty{100}{meters} until 2004 (see Figure \ref{fig:divts $a$}). Scaling this genus diversity by the cell count abundances, i.e. the Shannon index, shows a similar but less pronounced dynamic. Diversity remains high throughout 2000 and recovers more quickly but shows marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009, and 2016. The seasonal signal is lost (see Figure \ref{fig:divts} $f$).

@@

Create subsection like "Clustering analysis yields contrasting ecological regimes"
% Figure 4
We undertook clustering analysis on annually aggregated cell counts from microscopy and on individual monthly measurements. Figure \ref{fig:clustering}$a$ shows the results of clustering the annually aggregated communities using a linear complete linkage algorithm based on the binary Jaccard matrix. Two distinct clusters emerge from this analysis, with the first cluster encompassing the years 1996-2004 and 2014-2016, and the second cluster including the intermediate years from 2005 to 2013. 
Inter-annual species turnover is shown in Figure \ref{fig:clustering}$b$, it reveals an increasing trend in turnover from 1996 to 2004. Turnover subsequently decreases during the period underpinning cluster 2 and increases again in 2014, marking a return to cluster 1. This pattern is also reflected in the hierarchical clustering results, Figure \ref{fig:clustering}$a$, where cluster 2 shows a lower distance in the clustering tree between the years. We additionally extracted the three dominant genera for the second cluster and the early and late parts of the first cluster, respectively. The diatom \textit{Thalassiosira} dominated the early period and was abundant in all other periods. \textit{Pseudo-nitzschia} dominates cluster 2, with the haptophyte \textit{Emiliana} being the second most abundant genus. The later part of cluster 1 is dominated by \textit{Emiliana}.

@@

% P2: Discuss community reorganisation, biomass shifts vs HPLC data 
When comparing the two regimes, we observe a marked reduction in cell counts, which corresponds to a decreased concentration of chlorophyll \textit{a}, particularly in the top \qty{25}{meters} of the water column (see Figure \ref{fig:divts}). The decline in cell counts was not proportional to the more modest decrease in total chlorophyll \textit{a}, suggesting that the phytoplankton community might have transitioned to smaller cells, which were probably below the detection size limit of optical microscopes \cite{muller-karger_scientific_2019}. This shift to smaller cell sizes and a reduction in diatoms is consistent with predicted effects of warming and reduced mixing on phytoplankton communities \cite{bopp_response_2005}. Increases in temperature can lead to a decrease in the relative contribution of large cells in the community, irrespective of nutrient availability, but shifts towards smaller cells are particularly prevalent when temperatures increase and nutrient supply decreases \cite{mousing_global_2014}, as we observed in the Cariaco Basin.

>>>>

% Topic Overview
Phytoplankton are microscopic algae that form the foundation of the ocean food web. Because they respond quickly to environmental changes, studying them can show how marine ecosystems are affected by climate.
% Paper Overview
This study looked at how phytoplankton communities in the Cariaco Basin, off the coast of Venezuela, changed from 1995 to 2017. We compared microscopy cell counts and monthly water samples with weather data and two major climate indicators: the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both linked to local wind and currents.
% Findings Summary
The results showed a major shift in the phytoplankton community in 2004, linked to a decrease in upwelling, with conditions partially returning to the previous state by 2014. Interestingly, community diversity declined between 1998 and 2009, although upwelling had not yet changed. Further analysis showed that the AMO index, along with sea surface temperature, MEI v.2 and nitrate levels in the water column, were the strongest predictors of changes in the phytoplankton community.
% Key Takeaways
These findings contribute to a broader understanding of how climate variability shapes marine phytoplankton communities and underscore the importance of sustained long-term observations in detecting and interpreting ecological responses to climate change.

@@

\subsection{In-situ biological and environmental data}
    The CARIACO Ocean Time Series Program collected biological and environmental data, which we used here to examine the connections between environmental and ecological change in the region. Samples were collected on monthly research cruises at a station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.67}W) \cite{muller-karger_scientific_2019}.
    The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, along with other deployments and samples not utilized in this work. Details about sampling procedures, analytical methods, data quality control, and inter-calibration procedures have already been described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.

@@

    All in situ data, including the Niskin bottle \cite{mullerkarger2019niskin}, CTD \cite{mullerkarger2019ctd}, and phytoplankton taxonomy \cite{troccoli2019phytoplankton}, were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047}. 
    From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps (no more than six months) in coverage. In total, we used a complete set of 254 monthly data points from 1995 to 2017, for which all variables were available.

\subsection{Meteorological and climate data}
    %from Taylor et al. 2012
    In addition to oceanographic environmental data, we also examine meteorological and climatic data that are known to influence ocean conditions and marine ecosystems. Zonal wind (\textit{u}, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters \cite{rueda-roa_southern_2013}. We extracted the variable u10, the \qty{10}{\meter} eastward component of wind speed from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis \cite{hersbach2023era5} in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels}. For simplicity, we refer to this variable as "wind speed" throughout the article. The data for precipitation and evaporation are also extracted from the ERA5 dataset for the aforementioned grid cell. Evaporation provides an estimate of the accumulated amount of water that has evaporated from Earth's surface, while precipitation is an estimate of total precipitation, comprising all water that falls on Earth's surface as rain or snow.

    In addition to the wind data, we examined two climatic indices, each of which provides information on the basin-scale climatic conditions and, consequently, the broad-scale forcing of marine ecosystems. The monthly Atlantic Multidecadal Oscillation (AMO) index data (10-year low-pass), which represent area-averaged, detrended low-pass filtered North Atlantic SST anomalies (\ang{0}N-\ang{60}N) \cite{trenberth2023amo}, were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo}.

    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure, SLP; sea surface temperature, SST; zonal and meridional components of surface wind; and outgoing long-wave radiation, OLR) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W) \cite{noaa_psl_meiv2}, were obtained from \url{https://psl.noaa.gov/enso/mei/}. In our analysis, bi-monthly values were aligned with the tail month (December-January is treated as the value for January).

    \subsubsection{Data processing}
    % Niskin, CTD and Phytoplankton data processing
    Data were processed and analyzed using the R programming language (version 4.3.3) \cite{r_core_team_r_2024} with data pipelines created using tidyverse packages \cite{wickham_welcome_2019}. All the codes we produced for data processing and data analysis are publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.
    For all discrete depth measurements, the data were interpolated using the "oce" package \cite{kelley_oce_2023} with the adaptive "unesco" algorithm defined by the U.S. National Oceanographic Data Center \cite{johnson2006world}. The interpolated values were depth integrated to 100 meters and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 \meter for further analysis. Furthermore, the sea surface temperature (SST) was calculated from CTD temperature measurements averaged over the top \qty{10}{\meter}. For phytoplankton microscopy data, individually identified taxonomic units (species or genus) were interpolated and integrated across depth, and finally grouped by genus via the sum of interpolated counts.

\subsection{Data analysis}    
    In the analysis, we set out to find if there is a structure in the distribution and temporal evolution of diversity metrics and the community composition and to identify potential bottom-up drivers of this change. 
    To provide an overview of the available data and highlight interannual trends and changes in the time series data, we calculated annual anomalies for relevant variables as normalized z scores ($z = \frac{(x – \mu)}{\sigma}$, where $\mu$ is the mean across all available data points). Z scores from 1995, 2016, and 2017 were not shown because insufficient coverage (i.e. more than six missing monthly samples) in these years prevented us from producing reliable yearly estimates.

@@

    \subsubsection{Assessing Environmental Drivers of Phytoplankton Dynamics}
    To quantify the effects of bottom-up drivers in the Cariaco Basin on shifts in the phytoplankton community, we employed a Gradient Forest analysis. Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}. Gradient Forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor. The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community turnover. Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}. We used monthly genus level abundances integrated over the top \qty{100}{\meter} as indicator variables and limited the data to only those genera that were counted more than 10 times over the entire time series. This produced a data set of 193 monthly samples and 89 observed genera abundances, where all predictor variables were also available. The genus abundances were transformed by the natural logarithm plus one prior to model runs. The drivers/predictor variables considered in the analysis were the AMO and MEI v.2 indices, wind speed, evaporation, precipitation, \qty{21}{\celsius} isotherm, SST, salinity, and nutrient concentrations ($\mathrm{[NO_3]}$, $\mathrm{[PO_4]}$ and $\mathrm{[SiO_4]}$). All predictor variables are in some way related to the local climate and upwelling system and could be potential bottom-up drivers of community change, some more directly (e.g. the isotherm and nutrient concentrations) and some more indirectly, such as the climate indices and meteorological data. 
    The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology accounts for by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}. Gradient forest runs were performed with 1500 bootstrapped trees and a correlation threshold of 0.5.
    We show a model run without any time lags in the supplementary material (Figure \ref{fig:sup:GFoutput_nolags}). To consider the temporal lag in effect, particularly of climate indices, we tested for an improved predictive power of time lags of up to 6 months for all climate variables, as well as 3-month lags for in situ measurements in exploratory runs. We chose time lags that scored highest in the goodness-of-fit statistics for the final model run (Table \ref{tab:sup:GFlagtests} in Appendix). All the code and data used in this analysis and to create the plots are publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.

    \begin{table}
    \caption{Data and corresponding units used as predictor variables for the gradient forest analysis.}
    \centering
    \resizebox{\textwidth}{!}{\begin{tabular}[t]{rrl}
    \toprule
    Variable Name & Units  & Data source \\
    \midrule
    MEI v.2 & unitless & Climate index provided by NOAA \\
    AMO & unitless & Climate index provided by NSF NCAR \\
    Wind speed & \qty{}{\meter/\second} & ERA5 model reanalysis, negative 10 m u-component of wind speed \\
    Precipitation & \qty{}{\meter} & ERA5 model reanalysis, total precipitation\\
    Evaporation & \qty{}{\meter} & ERA5 model reanalysis, evaporation \\
    SST & \qty{}{\celsius} & CTD, Average temperature measured down to \qty{10}{\meter} depth\\
    \qty{21}{\celsius} isotherm & \qty{}{\meter} & CTD, depth at which temperature reaches 21°C\\
    Salinity & PSU & Niskin, interpolated down to \qty{100}{\meter} depth \\
    $\mathrm{NO}_3$  & µM & Niskin, interpolated down to \qty{100}{\meter} depth \\
    $\mathrm{PO}_4$  & µM & Niskin, interpolated down to \qty{100}{\meter} depth \\
    $\mathrm{SiO}_4$  & µM & Niskin, interpolated down to \qty{100}{\meter} depth \\
    \end{tabular}}
    \label{tab:DataSources}
    \end{table}

@@

\subsection{Interannual changes in the environment and ecosystem}
    % Figure 2
    Coordinated, interannual changes are observable in normalized yearly means of all environmental variables considered, the logarithmic transformation of chlorophyll \textit{a}, and phytoplankton cell counts per functional group (Figure \ref{fig:zscore}). The patterns emerging from wind speed, evaporation, precipitation, SST, and \qty{21}{\celsius} isotherm depth clearly show the years of positive anomalies of wind (blue) and negative anomalies for SST (red) and upwelling (red, i.e., shallower depth of the isotherm). These patterns are reflected in the mode of the AMO, where a positive anomaly correlates with weak upwelling regimes (see also Figure \ref{fig:sup:correlation} in Appendix). The yearly mean nutrient concentrations have a maximum positive anomaly at the beginning of the time series and are highly correlated. Annual mean anomalies of nitrate and phosphate are generally high during stronger upwelling, but remain relatively low between 2002 and 2014, with some variability in mean phosphate concentration. Silicate does not follow the pattern of nitrate or phosphate, but shows an elevated mean concentration between 2008 and 2013, which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion due to reduced abundances of diatoms.

    Coordinated, interannual changes are observable in normalized yearly means of all environmental variables considered, the logarithmic transformation of chlorophyll \textit{a}, and phytoplankton cell counts per functional group (Figure \ref{fig:zscore}). Distinct patterns emerge in wind speed, evaporation, precipitation, SST, and the \qty{21}{\celsius} isotherm depth. Periods of positive wind anomalies (blue) correspond to negative anomalies in SST and upwelling (red, indicated by a shallower isotherm depth), reflecting stronger upwelling conditions. These patterns are consistent with the Atlantic Multidecadal Oscillation (AMO), where a positive AMO phase is associated with weaker upwelling regimes (see also Figure~\ref{fig:sup:correlation} in the Appendix). Yearly mean nutrient concentrations show strong positive anomalies at the beginning of the time series and are generally well correlated. Nitrate and phosphate anomalies tend to be higher during periods of stronger upwelling but remain relatively low between 2002 and 2014, with nitrate showing more interannual variability. Silicate anomalies, in contrast, do not follow the same pattern as nitrate and phosphate. Instead, elevated concentrations are observed between 2008 and 2013, potentially driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced consumption due to lower diatom abundance.
    Chlorophyll \textit{a} and functional group cell counts display positive anomalies throughout most of the first half of the time series, followed by a significant reduction after 2004, suggesting a shift. This shift occurs during the period between 2003 and 2013, when the AMO index presents mostly positive anomalies, while upwelling-related variables indicate a weak upwelling regime (low wind, high SST, deep isotherm). Although data are sparse, nutrients and biological variables show a strong increase following 2013, coinciding with a return to negative anomalies in the AMO index and more upwelling-favorable conditions.
    The annual means of diversity metrics do not follow this trend, but are reduced between 1998 and 2008. Shannon diversity specifically appears elevated past 2006 and the general pattern that emerges is that mean diversity is lower in years, where mean chlorophyll \textit{a} is high. However, in 1996 and 1997 positive anomalies in cell counts, chlorophyll \textit{a} and diversity metrics coincide.

    \begin{figure}
    \noindent\includegraphics[width=\textwidth]{fig/Figure2_ZScores_v2.pdf}
    \caption{Time series showing annual z-scores of selected environmental, climatological, and ecological variables. Red indicates negative annual anomalies and blue indicates positive annual anomalies. Deeper shades indicate larger anomalies. Chlorophyll and functional group counts were log-transformed prior to normalisation. The years 1995 (n=2) and 2016 (n=5), 2017 (n=1) were removed due to insufficient monthly data to generate robust annual anomalies. Additionally we marked years with two or three and four or more missing monthly values.}
    \label{fig:zscore}
    \end{figure}

    % Figure 3
    Figure \ref{fig:divts} shows the depth distribution and dynamics of chlorophyll \textit{a} and community diversity indices, also in relation to rainy and upwelling seasons. The bulk of the chlorophyll \textit{a} ($\sim$ 80\,\%) was measured in the upper 50 meters of the water column, with approximately 15\,\% measured between 50 and 75 meters. Chlorophyll \textit{a} exhibited marked seasonality, with the highest biomass in the upper water column during the early upwelling season (December–February, Figure \ref{fig:divts}$b$). The yearly mean chlorophyll \textit{a} biomass showed the highest variability within the upper 25 meters. In some years, particularly in 2000, 2005, 2007, 2008, and 2013, the concentration in this layer was lower than in the 25–50 meter depth range, suggesting a deep chlorophyll maximum in those years (Figure \ref{fig:divts}$a$). The year 2013 was an outlier, particularly for the upper 25 meters, as missing monthly sampling from the upwelling season (January and February) skewed the value downwards. Nevertheless, the monthly measurements for the remainder of the year 2013 show a markedly lower cholorophyll (65\,\% of the mean over the entire time series for these months), which recovered in 2014. Chlorophyll \textit{a} biomass below 50 meters depth showed a relatively stable distribution throughout the year and across the time series. In fall, during the rainy season, mean chlorophyll \textit{a} biomass in the upper 25 meters was markedly reduced, while the 50–75 meter depth interval showed a slight increase, indicating a shift in the chlorophyll \textit{a} maximum to lower depths under weak upwelling conditions.

    The calculated genus diversity from the microscopy cell counts generally follows the dynamics of chlorophyll \textit{a} biomass, with maximum values observed during the upwelling seasons in winter. Genus richness is consistently the highest diversity index in the top \qty{25}{meters} depth interval, also showing the highest variability over the time series. Across the time series, we observe maximum richness at the start of the time series (1996-1997), with a subsequent reduction in diversity across all depths, with a recovery after 2011. The reduction in diversity following the start of the time series does not follow the pattern shown by chlorophyll \textit{a}, which remains relatively stable across the top \qty{100}{meters} until 2004 (see Figure \ref{fig:divts $a$}). Scaling genus richness by the cell count abundances, i.e. the Shannon index, shows a similar but less pronounced dynamic. Diversity remains high throughout 2000 and recovers more quickly but shows marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009, and 2016. The seasonal signal is lost (see Figure \ref{fig:divts} $f$).

@@

\subsection{Clustering analysis yields contrasting ecological regimes}
    % Figure 4
    We undertook clustering analysis on annually aggregated cell counts from microscopy and on individual monthly measurements. Figure \ref{fig:clustering}$a$ shows the results of clustering the annually aggregated communities using a linear complete linkage algorithm based on the binary Jaccard matrix. Two distinct clusters emerge from this analysis, with the first cluster encompassing the years 1996-2004 and 2014-2016, and the second cluster including the intermediate years from 2005 to 2013. 
    Inter-annual species turnover is shown in Figure \ref{fig:clustering}$b$, it reveals an increasing trend in turnover from 1996 to 2004. Turnover subsequently decreases during the period underpinning cluster 2 and increases again in 2014, marking a return to cluster 1. This pattern is also reflected in the hierarchical clustering results, Figure \ref{fig:clustering}$a$, where cluster 2 shows a lower distance in the clustering tree between the years. We additionally extracted the three dominant genera for the second cluster and the early and late parts of the first cluster, respectively. The diatom \textit{Thalassiosira} dominated the early period and was abundant in all other periods. \textit{Pseudo-nitzschia} dominates cluster 2, with the haptophyte \textit{Emiliana} being the second most abundant genus. The later part of cluster 1 is dominated by \textit{Emiliana}.

@@

    % P2: Discuss community reorganisation, biomass shifts vs HPLC data 
    When comparing the two clusters, we observe a marked reduction in cell counts, which corresponds to a decreased concentration of chlorophyll \textit{a}, particularly in the top \qty{25}{meters} of the water column (see Figure \ref{fig:divts}). The decline in cell counts was not proportional to the more modest decrease in total chlorophyll \textit{a}, suggesting that the phytoplankton community might have transitioned to smaller cells, which were probably below the detection size limit of optical microscopes \cite{muller-karger_scientific_2019}. This shift to smaller cell sizes and a reduction in diatoms is consistent with predicted effects of warming and reduced mixing on phytoplankton communities \cite{bopp_response_2005}. Increases in temperature can lead to a decrease in the relative contribution of large cells in the community, irrespective of nutrient availability, but shifts towards smaller cells are particularly prevalent when temperatures increase and nutrient supply decreases \cite{mousing_global_2014}, as we observed in the Cariaco Basin.
